### 2024-07-08 latex/HP1_comm.tex
new input files

In his commentary {\emph Jyotsnā} on \emph{Haṭhapradīpikā} 1.1 Brahmānanda first states that the
author Svātmārāma starts appropriately with a \emph{maṅgala}, a verse of adoration addressing
Ādinātha, i.e.\ Śiva, but in accordance with his non-sectarian approach leaves room for a second
interpretation of the word as Viṣṇu. A sentence later this is contradicted by his explanantion that Ādinātha, who first taught Yoga, taught it to Pārvatī, which limits the scope to Śiva.

%add comment about yoga replacing mārga and saudha here and in the next two verses.
The reading \emph{°rājasaudham} in the third verse quarter is well attested by the manuscripts, including \alphaOne. However, the most common reading in manuscripts on the lower branches of the stemma is \emph{°rājayogam}, which was accepted by Brahmānanda in \emph{Jyotsnā} 1.1 (see below). It appears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.

@@

\begin{versinnote}
śrīādināthāya namo 'stu tasmai\\ 
\ \ yenopadiṣṭā haṭhayogavidyā |\\
vibhrājate pronnatarājayogam\\ 
\ \ āroḍhum icchor adhirohiṇīva ||1.1||
\end{versinnote}

Our choice of \emph{virājate} in the third verse quarter reflects the relevant manuscripts and rhymes with the following \emph{rāja-}.  
The difference in meaning is negligable.

The main problem in Brahmānanda's interpretation is his choice of \emph{-rājayogaṃ} over
\emph{-rājasaudhaṃ}, which he explains as follows:

\begin{quote}
  \emph{rājayogaś ca sarvavṛttini\-rodhalakṣaṇo 'samprajñātayogaḥ. tam icchor mumukṣor adhirohiṇīva
  adhi\-ruhyate 'nayety adhirohiṇī niḥśreṇīva vibhrājate viśeṣeṇa bhrājate śobhate. yathā
  pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudha\-prāpi\-kā bhavati evaṃ
  haṭhapradīpikāpi pronnatarājayogam āroḍhum icchor anā\-yāsena rājayogaprāpikā bhavatīti
  upamālaṅkāraḥ.  indravajrākhyaṃ vṛttam.}

@@

The interpretation makes good sense: Haṭhayoga leads effortlessly to Rājayoga, just as a ladder
leads one to the flat on the top floor. And the comparison as outlined by Brahmānanda has all the
elements deemed necessary by Sanskritic poetology:

\begin{enumerate}
\item Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva} 
\item Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā
    adhirohiṇī}) 
\item Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The mūla text has
    \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā
    haṭhapradīpikā}) 
\item Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})
\end{enumerate}

@@

@@

%%%%%%%%%%
\subsection*{1.35}
\begin{translation}[hp01_035a]
Now, the adept's pose (\emph{siddhāsana}).
\end{translation}

\begin{translation}[hp01_035]
[The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, focus the mind, hold the body erect and [remain] motionless, his senses restrained, gazing between the brows with his eyes unmoving. This, which breaks open the door to liberation, is called the adept’s pose.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.36}
\begin{translation}[hp01_036]
However, in another school [\emph{siddhāsana} is taught as follows]:\\

Place the left heel on the penis and put the other heel on top: this is the adept's pose (\emph{siddhāsana}).\\

Only the first teaching [on \emph{siddhāsana}] is accepted by me.\\

@@

\begin{philcomm}[hp01_036]
Svātmārāma's introductory and following remarks to verse 1.36 indicate that he preferred the \emph{siddhāsana} of the \emph{Vivekamārtaṇḍa} over the version taught as \emph{muktāsana} in the \emph{Vasiṣṭhasaṃhitā} and \emph{Yogayājñavalkya}.
\end{philcomm}
%%%%%%%%%%
\subsection*{1.37}
\begin{translation}[hp01_037]
Some call this the adept's pose (\emph{siddhāsana}), others know it as the thunderbolt pose (\emph{vajrāsana}), a few say it is the pose of the liberated (\emph{muktāsana}) and some call it the secret pose (\emph{guptāsana}).
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.44}
\begin{translation}[hp01_044]
Now the lotus pose (\emph{padmāsana}).\\

Place the right foot on the left thigh, and the left on the right thigh, firmly hold the big toes with the hands crossed behind the back, put the chin on the chest and gaze at the tip of the nose. This, which destroys diseases for those who undertake the observances, is called the lotus pose.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.45--46}
\begin{translation}[hp01_045]
However, in another school [\emph{padmāsana} is taught as follows]:\\

Carefully put the upturned feet on the thighs and the upturned hands in the middle of the thighs, fix the eyes on the tip of the nose, raise the root of the uvula with the tongue, place the chin on the chest, gently [draw in] the breath [...].
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.47}
\begin{translation}[hp01_047]
This is called the lotus pose [and] it cures all diseases. It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.\\

Only the second teaching [on the lotus pose] is approved by me.  
\end{translation}

@@

\begin{philcomm}[hp01_047]
In this context, the word \emph{durlabham} is somewhat ambiguous as to whether the posture is hard to perform or hard to acquire (the more usual meaning). In commenting on \emph{durlabham} in \emph{Jyotsnā} 2.74, Brahmānanda glosses it as \emph{duṣprāpam}, which means `difficult to attain' and `inaccessible.'

The comment added to this verse by Svātmārāma indicates that he prefers the following version of \emph{padmāsana}, which derives from the \emph{Vivekamārtaṇḍa}, rather than the one he has borrowed from the \emph{Dattātreyayogāśāstra}. 
% Maybe add some remarks by commentators on how they understand it.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.50--52}
\begin{translation}[hp01_050]
Now, the lion’s pose (\emph{siṃhāsana}).

[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. He should place the left ankle on right, the right ankle on the left and both hands on the knees, spread his fingers, open his mouth and gaze in deep concentration at the tip of his nose. This is the lion’s pose, which is always honoured by yogis. It causes the three locks to arise together and is the best of [all] postures.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.53--54}
\begin{translation}[hp01_053]
Now, the friendly pose (\emph{bhadrāsana}).

[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.  By firmly and very steadily holding the sides of the feet with the hands, the friendly pose arises, which cures all diseases and poisons. Yogis of the Siddha tradition call it Gorakṣa's pose (\emph{gorakṣāsana}).
\end{translation}

@@

@@

%%%%%%%%%%
\subsection*{1.61}
\begin{translation}[hp01_061]
In the same vein there is a saying by Goraksa:
One should avoid places near bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.
\end{translation}

>>>>

In his commentary \emph{Jyotsnā} on \emph{Haṭhapradīpikā} 1.1 Brahmānanda first states that the
author Svātmārāma starts appropriately with a \emph{maṅgala}, a verse of adoration addressing
Ādinātha, i.e.\ Śiva, but in accordance with his non-sectarian approach leaves room for a second
interpretation of the word as Viṣṇu. A sentence later this is contradicted by his explanantion that Ādinātha, who first taught Yoga, taught it to Pārvatī, which limits the scope to Śiva.\lb

%add comment about yoga replacing mārga and saudha here and in the next two verses.
The reading \emph{°rājasaudham} in the third verse quarter is well attested by the manuscripts, including \alphaOne. However, the most common reading in manuscripts on the lower branches of the stemma is \emph{°rājayogam}, which was accepted by Brahmānanda in \emph{Jyotsnā} 1.1 (see below). It appears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.\lb

@@

\begin{versinnote}
\tl{śrīādināthāya namo 'stu tasmai\\+} 
\tl{\ \ yenopadiṣṭā haṭhayogavidyā\\+}
\tl{vibhrājate pronnatarājayogam\\+}
\tl{\ \ āroḍhum icchor adhirohiṇīva ||1.1||}
\end{versinnote}

Our choice of \emph{virājate} in the third verse quarter reflects the relevant manuscripts and rhymes with the following \emph{rāja-}.  
The difference in meaning is negligable.\lb

The main problem in Brahmānanda's interpretation is his choice of \emph{-rājayogaṃ} over
\emph{-rājasaudhaṃ}, which he explains as follows:\lb

\begin{quote}
  \emph{rājayogaś ca sarvavṛttini\-rodhalakṣaṇo 'samprajñātayogaḥ. tam icchor mumukṣor adhirohiṇīva
  adhi\-ruhyate 'nayety adhirohiṇī niḥśreṇīva vibhrājate viśeṣeṇa bhrājate śobhate. yathā
  pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudha\-prāpi\-kā bhavati evaṃ
  haṭhapradīpikāpi pronnatarājayogam āroḍhum icchor anā\-yāsena rājayogaprāpikā bhavatīti
  upamālaṅkāraḥ.  indravajrākhyaṃ vṛttam.}\lb

@@

The interpretation makes good sense: Haṭhayoga leads effortlessly to Rājayoga, just as a ladder
leads one to the flat on the top floor. And the comparison as outlined by Brahmānanda has all the
elements deemed necessary by Sanskritic poetology:\lb

\begin{enumerate}
\item{Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva}}
\item{Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā adhirohiṇī})}
\item{Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The mūla text has \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā haṭhapradīpikā})}
\item{Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})}
\end{enumerate}

@@

%%%%%%%%%%
% \subsection*{1.17 heading}
% athāsanāni/
% \begin{translation}[hp01_017a]
% \end{translation}
% \begin{philcomm}[hp01_017a]
% \end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.35 heading}
\begin{translation}[hp01_035a]
Now, the adept's pose (\emph{siddhāsana}).
\end{translation}

% \begin{philcomm}[hp01_035a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{1.35}
\begin{translation}[hp01_035]
[The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, focus the mind, hold the body erect and [remain] motionless, his senses restrained, gazing between the brows with his eyes unmoving. This, which breaks open the door to liberation, is called the adept’s pose.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.36 heading}
\begin{translation}[hp01_036a]
However, in another school [\emph{siddhāsana} is taught as follows]:
\end{translation}

% \begin{philcomm}[hp01_036a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{1.36}
\begin{translation}[hp01_036]
Place the left heel on the penis and put the other heel on top: this is the adept's pose (\emph{siddhāsana}).
\end{translation}

@@

\begin{philcomm}[hp01_036]
Svātmārāma's introductory and following remarks to verse 1.36 indicate that he preferred the \emph{siddhāsana} of the \emph{Vivekamārtaṇḍa} over the version taught as \emph{muktāsana} in the \emph{Vasiṣṭhasaṃhitā} and \emph{Yogayājñavalkya}.
\end{philcomm}
%%%%%%%%%%
\subsection*{1.36 ending}
\begin{translation}[hp01_036p]
Only the first teaching [on \emph{siddhāsana}] is accepted by me.
\end{translation}

% \begin{philcomm}[hp01_036p]
% \end{philcomm}

%%%%%%%%%%
\subsection*{1.37}
\begin{translation}[hp01_037]
Some call this the adept's pose (\emph{siddhāsana}), others know it as the thunderbolt pose (\emph{vajrāsana}), a few say it is the pose of the liberated (\emph{muktāsana}) and some call it the secret pose (\emph{guptāsana}).
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.44 heading}
\begin{translation}[hp01_044a]
Now the lotus pose (\emph{padmāsana}).
\end{translation}

% \begin{philcomm}[hp01_044a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{1.44}
\begin{translation}[hp01_044]
Place the right foot on the left thigh, and the left on the right thigh, firmly hold the big toes with the hands crossed behind the back, put the chin on the chest and gaze at the tip of the nose. This, which destroys diseases for those who undertake the observances, is called the lotus pose.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.45 heading}
\begin{translation}[hp01_045a]
However, in another school [\emph{padmāsana} is taught as follows]:
\end{translation}

% \begin{philcomm}[hp01_045a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{1.45--46}
\begin{translation}[hp01_045]
Carefully put the upturned feet on the thighs and the upturned hands in the middle of the thighs, fix the eyes on the tip of the nose, raise the root of the uvula with the tongue, place the chin on the chest, gently [draw in] the breath [...].
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.47}
\begin{translation}[hp01_047]
This is called the lotus pose [and] it cures all diseases. It is difficult for just anyone to accomplish; it is accomplished by a wise person [here] on earth.
\end{translation}

@@

\begin{philcomm}[hp01_047]
In this context, the word \emph{durlabham} is somewhat ambiguous as to whether the posture is hard to perform or hard to acquire (the more usual meaning). In commenting on \emph{durlabham} in \emph{Jyotsnā} 2.74, Brahmānanda glosses it as \emph{duṣprāpam}, which means `difficult to attain' and `inaccessible.' 
\end{philcomm}

%%%%%%%%%%
\subsection*{1.47 ending}
\begin{translation}[hp01_047p]
Only the second teaching [on the lotus pose] is approved by me.  
\end{translation}

\begin{philcomm}[hp01_047p]
The comment added to this verse by Svātmārāma indicates that he prefers the following version of \emph{padmāsana}, which derives from the \emph{Vivekamārtaṇḍa}, rather than the one he has borrowed from the \emph{Dattātreyayogāśāstra}. 
% Maybe add some remarks by commentators on how they understand it.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.50 heading}
\begin{translation}[hp01_050a]
Now, the lion’s pose (\emph{siṃhāsana}).
\end{translation}

% \begin{philcomm}[hp01_050a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{1.50--52}
\begin{translation}[hp01_050]
[The yogi] should put both ankles at the sides of the perineal seam below the scrotum. He should place the left ankle on right, the right ankle on the left and both hands on the knees, spread his fingers, open his mouth and gaze in deep concentration at the tip of his nose. This is the lion’s pose, which is always honoured by yogis. It causes the three locks to arise together and is the best of [all] postures.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.53 heading}
\begin{translation}[hp01_053a]
Now, the friendly pose (\emph{bhadrāsana}).
\end{translation}

% \begin{philcomm}[hp01_053a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{1.53--54}
\begin{translation}[hp01_053]
[The yogi] should put both ankles at the sides of the perineal seam below the scrotum.  By firmly and very steadily holding the sides of the feet with the hands, the friendly pose arises, which cures all diseases and poisons. Yogis of the Siddha tradition call it Gorakṣa's pose (\emph{gorakṣāsana}).
\end{translation}

@@

%%%%%%%%%%
% \subsection*{1.54 endding}
% \begin{translation}[hp01_054p]
% % ity āsanāni
% \end{translation}
% \begin{philcomm}[hp01_054p]
% \end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.61 heading}
\begin{translation}[hp01_061a]
In the same vein there is a saying by Goraksa:
\end{translation}

% \begin{philcomm}[hp01_061a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{1.61}
\begin{translation}[hp01_061]
One should avoid places near bad people, frequenting fire, women and roads, and observances which harm the body such as early morning bathing and fasting.
\end{translation}

### 2024-07-08 latex/HP2_comm.tex
new input files

%%%%%%%%%%
\subsection*{2.24}
\begin{translation}[hp02_024]
Among them is dhauti:\\ 
{}[The yogi] should slowly swallow a moistened cloth  four finger-breadths in width and then draw it out. This ejection [of it] from the mouth is the dhauti technique.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.26}
\begin{translation}[hp02_026]
Now the \emph{basti} technique.\\
Squatting in water up to the navel with a reed inserted in the anus, [the yogi] should contract the perineal region (\emph{ādhāra°}). The [resultant] flushing is the basti technique.
\end{translation}
% Adopt pakhālaṃ — not done in latest version

@@

%%%%%%%%%%
\subsection*{2.30}
\begin{translation}[hp02_030]
Now \emph{neti}.\\
{[}The yogi] should insert a very smooth thread one handspan [in length] into the nasal passage and take it out through the mouth. This is called neti by the Siddhas.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.32}
\begin{translation}[hp02_032]
Now \emph{trāṭaka}:\\
{}[The yogi] should concentrate and look at a small focal point with a fixed gaze until tears fall. The experts consider this to be trāṭaka.
\end{translation}
% adopt matam

@@

%%%%%%%%%%
\subsection*{2.34}
\begin{translation}[hp02_034]
Now \emph{naulī}:\\
With the shoulders lowered, [the yogi] should rotate the stomach to the left and right with the speed of a rapid whirlpool. This is called naulī by people from Gauḍa.%
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.36}
\begin{translation}[hp02_036]
Now kapālabhātī:\\
Rapid inhalation and exhalation like the bellows of a blacksmith is called kapālabhātī, the skull bellows. It dries up imbalances of phlegm.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.48}
\begin{translation}[hp02_048]
Now, piercing the sun---\\
The yogi should sit in \emph{vajrāsana} on a comfortable mat, slowly draw in external air through the right nostril, [...]
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.51}
\begin{translation}[hp02_051]
Now ujjāyī:

[The yogi] should close the mouth and gradually draw in the breath through the nostrils so that it comes into contact [with the region] from the throat to the chest and makes a sound.\end{translation}

@@

%%%%%%%%%%
\subsection*{2.54}
\begin{translation}[hp02_054]
Now sītkā---

{}[The yogi] should continuously make \emph{sīt} sound in the mouth and flare his nostrils. By practising in this way he becomes a second god of love.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.57}
\begin{translation}[hp02_057]
And the very same has been taught [as follows]:

He who continuously takes in the breath through the tongue and the root of the palate has all his diseases cured in half a year.
\end{translation}

@@

\begin{philcomm}[hp02_057]
Verse 2.57 seems to be describing an alternative method of \emph{sītkā kumbhaka}. The introductory phrase \emph{evam eva uktaṃ ca} suggests that the teaching in this verse is consistent with what preceded it. However, the first version of \emph{sītkā} appears to be done at the front of the mouth (\emph{vaktra}), whereas the next version is done at the back of the mouth (see below).%JM Not sure what the note is saying. This practice could be consistent with 2.54 although no mention is made of a sound.

@@

%%%%%%%%%%
\subsection*{2.58}
\begin{translation}[hp02_058]
Now śītalī:\\
The wise man should draw in air through the tongue and after retaining the breath as before gradually exhale through the nostrils .
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.60}
\begin{translation}[hp02_060]
Now bhastrikā:

If [the yogi] places the soles of both feet on the thighs, the lotus pose, which destroys all bad deeds, duly arises.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.69}
\begin{translation}[hp02_069]
Now bhrāmarī:

Forcibly loud inhalation with the sound of a male bee; very slow exhalation with the sound of a female bee: as a result of practising thus, there arises in the minds of the best yogis a certain blissful stupefaction.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.70}
\begin{translation}[hp02_070]
Now mūrchā:

At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This fainting of the mind, which is called \emph{mūrchanā}, gives pleasure.
\end{translation}
%name is given as mūrchanā but headings in alpha etc. are mūrchā. Some, including epsilon1 do have mūrchanā.
% adopt feminine in 2nd hemistich
% JB: the meaning of manomūrcchā is not conveyed by the current translation. Fainting of the mind is different to normal fainting where a person falls to the ground unconsciousness. It's unlikely that the person in mūrcchā kumbhaka was supposed to collapse.... bestows bliss, when the mind faints.

@@

%%%%%%%%%%
\subsection*{2.71}
\begin{translation}[hp02_071]
Now plāvanī:\\
With his abdomen completely filled with the wind of eructation, which has been turned inwards, [the yogi] floats easily even in deep water, like a lotus leaf.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.72}
\begin{translation}[hp02_072]
Now \emph{kevalakumbhaka}:

>>>>

%%%%%%%%%%
\subsection*{2.24 heading}
\begin{translation}[hp02_024a]
Among them is dhauti:
\end{translation}

% \begin{philcomm}[hp02_024a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.24}
\begin{translation}[hp02_024]
{}[The yogi] should slowly swallow a moistened cloth  four finger-breadths in width and then draw it out. This ejection [of it] from the mouth is the dhauti technique.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.26 heading}
\begin{translation}[hp02_026a]
Now the \emph{basti} technique.
\end{translation}

% \begin{philcomm}[hp02_026a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.26}
\begin{translation}[hp02_026]
Squatting in water up to the navel with a reed inserted in the anus, [the yogi] should contract the perineal region (\emph{ādhāra°}). The [resultant] flushing is the basti technique.
\end{translation}
% Adopt pakhālaṃ — not done in latest version

@@

%%%%%%%%%%
\subsection*{2.30 heading}
\begin{translation}[hp02_030a]
Now \emph{neti}.
\end{translation}

% \begin{philcomm}[hp02_030a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.30}
\begin{translation}[hp02_030]
{[}The yogi] should insert a very smooth thread one handspan [in length] into the nasal passage and take it out through the mouth. This is called neti by the Siddhas.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.32 heading}
\begin{translation}[hp02_032a]
Now \emph{trāṭaka}:
\end{translation}

% \begin{philcomm}[hp02_032a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.32}
\begin{translation}[hp02_032]
{}[The yogi] should concentrate and look at a small focal point with a fixed gaze until tears fall. The experts consider this to be trāṭaka.
\end{translation}
% adopt matam

@@

%%%%%%%%%%
\subsection*{2.34 heading}
\begin{translation}[hp02_034a]
Now \emph{naulī}:
\end{translation}

% \begin{philcomm}[hp02_034a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.34}
\begin{translation}[hp02_034]
With the shoulders lowered, [the yogi] should rotate the stomach to the left and right with the speed of a rapid whirlpool. This is called naulī by people from Gauḍa.%
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.36 heading}
\begin{translation}[hp02_036a]
Now kapālabhātī:
\end{translation}

% \begin{philcomm}[hp02_036a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.36}
\begin{translation}[hp02_036]
Rapid inhalation and exhalation like the bellows of a blacksmith is called kapālabhātī, the skull bellows. It dries up imbalances of phlegm.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.48 heading}
\begin{translation}[hp02_048a]
Now, piercing the sun:
\end{translation}

% \begin{philcomm}[hp02_048a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.48}
\begin{translation}[hp02_048]
The yogi should sit in \emph{vajrāsana} on a comfortable mat, slowly draw in external air through the right nostril, [...]
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.51 heading}
\begin{translation}[hp02_051a]
Now ujjāyī:
\end{translation}

% \begin{philcomm}[hp02_051a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.51}
\begin{translation}[hp02_051]
[The yogi] should close the mouth and gradually draw in the breath through the nostrils so that it comes into contact [with the region] from the throat to the chest and makes a sound.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.54 heading}
\begin{translation}[hp02_054a]
Now sītkā:
\end{translation}

% \begin{philcomm}[hp02_054a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.54}
\begin{translation}[hp02_054]
{}[The yogi] should continuously make \emph{sīt} sound in the mouth and flare his nostrils. By practising in this way he becomes a second god of love.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.57 heading}
\begin{translation}[hp02_057a]
And the very same has been taught [as follows]:
\end{translation}

% \begin{philcomm}[hp02_057a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.57}
\begin{translation}[hp02_057]
He who continuously takes in the breath through the tongue and the root of the palate has all his diseases cured in half a year.
\end{translation}

@@

\begin{philcomm}[hp02_057]

@@

%%%%%%%%%%
\subsection*{2.58 heading}
\begin{translation}[hp02_058a]
Now śītalī:
\end{translation}

% \begin{philcomm}[hp02_058a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.58}
\begin{translation}[hp02_058]
The wise man should draw in air through the tongue and after retaining the breath as before gradually exhale through the nostrils .
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.60 heading}
\begin{translation}[hp02_060a]
Now bhastrikā:
\end{translation}

% \begin{philcomm}[hp02_060a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.60}
\begin{translation}[hp02_060]
If [the yogi] places the soles of both feet on the thighs, the lotus pose, which destroys all bad deeds, duly arises.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.69 heading}
\begin{translation}[hp02_069a]
Now bhrāmarī:
\end{translation}

% \begin{philcomm}[hp02_069a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.69}
\begin{translation}[hp02_069]
Forcibly loud inhalation with the sound of a male bee; very slow exhalation with the sound of a female bee: as a result of practising thus, there arises in the minds of the best yogis a certain blissful stupefaction.
\end{translation}

@@

%%%%%%%%%%
\subsection*{2.70 heading}
\begin{translation}[hp02_070a]
Now mūrchā:
\end{translation}

% \begin{philcomm}[hp02_070a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.70}
\begin{translation}[hp02_070]
At the end of inhalation [the yogi] should tightly apply the Jālandhara [lock] and exhale slowly. This fainting of the mind, which is called \emph{mūrchanā}, gives pleasure.
\end{translation}
%name is given as mūrchanā but headings in alpha etc. are mūrchā. Some, including epsilon1 do have mūrchanā.
% adopt feminine in 2nd hemistich
% JB: the meaning of manomūrcchā is not conveyed by the current translation. Fainting of the mind is different to normal fainting where a person falls to the ground unconsciousness. It's unlikely that the person in mūrcchā kumbhaka was supposed to collapse.... bestows bliss, when the mind faints.

@@

%%%%%%%%%%
\subsection*{2.71 heading}
\begin{translation}[hp02_071a]
Now plāvanī:
\end{translation}

% \begin{philcomm}[hp02_071a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.71}
\begin{translation}[hp02_071]
With his abdomen completely filled with the wind of eructation, which has been turned inwards, [the yogi] floats easily even in deep water, like a lotus leaf.
\end{translation}

@@

%%%%%%%%%%
% \subsection*{2.71 ending}
% iti plāvanīkumbhakam/
% \begin{translation}[hp02_071p]
% \end{translation}
% \begin{philcomm}[hp02_071p]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.72 heading}
\begin{translation}[hp02_072a]
Now \emph{kevalakumbhaka}:
\end{translation}

% \begin{philcomm}[hp02_072a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{2.72}
\begin{translation}[hp02_072]
